---
title: "R scripts and R markdown files"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

R code can be incorporated in R markdown files. This is a great way to get data and code to work together with the report. However, not all code is suitable to run everytime you want to compile (or **knit**) your report. This is beacause some code is quite heavy and will take time to evaluate. Another case is when the code is quite long. You might want to keep the code in another file to save space in your markdown file. 

## Write and store R scripts
R Scripts can be thought of as small programs. It is good practice that each script can be executed on its own. This means that the script should contain information on what libraries and what data to use. The script (or program) will perform all computations that you tell it, in the order that you specify. 

When writing code, remember to write what you intend to do. This is actually a good way to start each script. Use `#` to add comments to your script. Each line starting with a `#` will not be evaluated by R.

Here is an example script, whitout the code:

```{r, eval = FALSE}

# Purpose: Calculate the mean of 10 random numbers and store the result in the environment


# Create random numbers  ############


# Calculate the mean and store in environment ###########


# End script ################

```

I have written a plan. The script starts with a purpose, this can be even more detailed so that when you continue to work on this script later, you know what you wanted to do. 
Next line contains a statement what the code below should do. The line ends with multiple `#`, this makes it easy to find this part later as R Studio will interpret it as a headline and you can find it in a menu below the script pane.

Let's add some code to the plan above.

```{r, eval = FALSE}

# Purpose: Calculate the mean of 10 random numbers and store the result in the environment


# Create random numbers  ############

# rnorm() creates random numbers based on the mean and standard deviation

numbers <- rnorm(10, 100, 10) # n = 10, mean = 100, sd = 10

# Calculate the mean and store in environment ###########

m <- mean(number) # store the mean in object m


# End script ################

```

This is of course a silly example as we create some fake data from a distribution with a known mean and standard deviation. It will differ slightly from 100 because of the process of generating the ten number. Anyway, the number is stored in an object called `m`. The script has done it's job. 

Let's say that we save the script as `./R/calculate_mean.R`. And we want to get the variable `m` into our environment from anouther script or R markdown file. We can use the `source()` function to do this. `source()` will run the script and any calculations or functions specified in the script will be executed. Remember that the script should be self contained, meaining that all information on running it should be specified in the script it selfe. This makes it possible to source script files. 

An example of using source to run another script:

```{r, eval = FALSE}

source("./R/calculate_mean.R") # Calculate mean as specified in script

```

## R Markdown files
Think of R Markdown files as special programs used to execute operations together with text. The result from a R Markdown file can also contain figures and tables created from data. The same principles apply here as when writing R scripts. Use a lot of comments to make it clear what you intend to do. As covered earlier, R code is included as code chunks. These can be commented using the `#` symbol. When commenting outside code chunks  one can use html-comments, these comes in the form ```<!-- this is comment -->```.

# Conclusions

- Use relative paths and self-contained scripts.
- Use comments to describe what the purpose of your script is and what different functions do.
- Store scripts with good, descriptive names.  



